Abstractions for Type-Level Programming

### 2022-04-07 article.tex
Add Richard's comments
Define Λ and V

\FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$ and class disjointness $\Ξ$. \hl{\text{Highlights}} correspond to additions to \SystemFsub.}

@@

\subsubsection*{Disjointness / Subtyping Exclusivity}
The following non-standard lemma is necessary to prevent overlap between the \SMatch1/2 and \SMatch3/4 rules.
%
\input{proofs/disjointness-subtyping-exclusivity.tex}
%
If such overlap would be allowed, match types could reduce in several different ways, resulting in an unsound system.
We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$.
Elements of $\P$ can be understood as equivalence classes for \Fm's types.
We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
This set-theoretical view lets us conclude the desired result directly.
In our Coq mechanization, we axiomatize this lemma and delegate to the pen-and-paper proof.

@@

The following \Cref{lem:inversion-of-subtyping} allows us to perform inversion on the subtyping relation,
which is important to show canonical forms (\Cref{lem:canonical-forms}) and inversion of typing (\Cref{lem:inversion-of-typing}).
Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
It represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
%
\input{proofs/inversion-of-subtyping.tex}
%
The first point of \Cref{lem:inversion-of-subtyping} uses the structure of the $\⇌$ to provide a form of inversion, which we use to prove each of the subsequent points.
In comparison with the corresponding inversion lemma in \Fsub's safety proof, the statement and the proof of \Cref{lem:inversion-of-subtyping} are longer and more intricate.
This difference is inevitable, given that match type reduction rules allow match expressions to be typed as the result of their reduction, which complexifies the inversion.

@@

In \SystemFmB (\Cref{subsec:type-binding-extension}), match expressions support two sorts of parametric patterns: they can be either instantiated (\!\!$\match \{ \xs \: \List~\Int \}$, where $\Int \∈ \A$), or use a binding pattern variable (\!\!$\match \[\X] \{ \xs \: \List~\X \}$, where $\X$ is a pattern variable).
In this sense, \SystemFmB is \emph{more expressive} than Scala, where instantiated patterns are not available at the term level due to type erasure.

@@

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, sjoberg2015a, eisenberg2016dependent}.

@@

\SystemFm's type rules enable any match expression to be typed as a match type, but the situation is different in the full Scala language.
Pattern matching in Scala supports many sorts of patterns~\citep{emir2007matching}, most of which do not have a match type counterpart.
Furthermore, typing match expressions as match types is not enabled by default in order to preserve backward compatibility.
Instead, explicit type annotations must be provided.

>>>>

% TODO: define Ψ and Ξ
% TODO: cite the precise figure in Pierce 2002
\FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$ and class disjointness $\Ξ$. \hl{\text{Highlights}} correspond to additions to \SystemFsub.}

@@

\subsubsection*{Disjointness / Subtyping Exclusivity}
The following non-standard lemma is necessary to prevent overlap between the \SMatch1/2 and \SMatch3/4 rules.
%
\input{proofs/disjointness-subtyping-exclusivity.tex}
%
If such overlap would be allowed, match types could reduce in several different ways, resulting in an unsound system.
We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representative for abstractions and type abstractions, respectively.
Elements of $\P$ can be understood as equivalence classes for \Fm's types.
We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
This set-theoretical view lets us conclude the desired result directly.
In our Coq mechanization, we axiomatize this lemma and delegate to the pen-and-paper proof.

@@

The following \Cref{lem:inversion-of-subtyping} allows us to perform inversion on the subtyping relation,
which is important to show canonical forms (\Cref{lem:canonical-forms}) and inversion of typing (\Cref{lem:inversion-of-typing}).
Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
% TODO: clarify that harpoon is asymmetric (and so it disj)
It represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
%
\input{proofs/inversion-of-subtyping.tex}
%
The first point of \Cref{lem:inversion-of-subtyping} uses the structure of the $\⇌$ to provide a form of inversion, which we use to prove each of the subsequent points.
In comparison with the corresponding inversion lemma in \Fsub's safety proof, the statement and the proof of \Cref{lem:inversion-of-subtyping} are longer and more intricate.
This difference is inevitable, given that match type reduction rules allow match expressions to be typed as the result of their reduction, which complexifies the inversion.

@@

In \SystemFmB (\Cref{subsec:type-binding-extension}), match expressions support two sorts of parametric patterns: they can be either instantiated (\!\!$\match \{ \xs \: \List~\Int \}$, where $\Int \∈ \A$), or use a binding pattern variable (\!\!$\match \[\X] \{ \xs \: \List~\X \}$, where $\X$ is a pattern variable).
In this sense, \SystemFmB is \emph{more expressive} than Scala, where instantiated patterns are not available at the term level due to type erasure.
% TODO: say that our approach would not work in a system with full type erasure

@@

Unlike our calculus, the Scala implementation also allows match type definitions to be recursive.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, sjoberg2015a, eisenberg2016dependent}.

@@

\SystemFm's type rules enable any match expression to be typed as a match type, but the situation is different in the full Scala language.
Pattern matching in Scala supports many sorts of patterns~\citep{emir2007matching}, most of which do not have a match type counterpart.
Furthermore, typing match expressions as match types is not enabled by default in order to preserve backward compatibility.
Instead, explicit type annotations must be provided.
% TODO: Add examples
